---
title: "Postcode to BRMA lookup - v1"
author: "Nick Bailey"
date: '2023-05-25'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# # first time only
# install.packages("here")
# install.packages("sf")
# install.packages("tidyverse")

library(here)
library(sf)
library(tidyverse)

```


# Constructing a postcode to BRMA lookup file

This Rmd creates a lookup file linking unit postcodes to Broad Rental Market Area (BRMA) boundaries for Great Britain. BRMAs are the spatial areas used for the administration of Housing Benefit or Local Housing Allowance in GB. There are 192 BRMAs in total.

Final lookup ('pcode_brma_lookup.csv') has the following fields: 

* PCD: 7-character postcodes (may have one space or none)
* PCD2: 8-character postcodes (may have one space or two)
* PCDS: 7- or 8-character postcodes (always has one space)
* brma: the reference number, taken from the BRMA boundary files - see details below
* brma_name: the name for the BRMA.

## BRMA boundaries 

### England
April 2014 version: Boswarva, Owen. (2017). Broad Rental Market Areas (England), [Dataset]. independent researcher. https://doi.org/10.7488/ds/1963. Obtained using FOI. Open Government Licence.
Dataset released by the Valuation Office Agency on 16 September 2014. 

There is another version from May 2020 (https://www.gov.uk/government/publications/broad-rental-market-area-boundary-layer-for-geographical-information-system-gis-applicable-may-2020) but number of BRMAs and size of BRMAs (very near) identical in all cases. 

```{r brma eng}

# convert to BNG
sf_brma_eng <- st_read(here("data-input", "brma_eng", "BRMA0412.shp"))
# st_crs(sf_brma_eng)
sf_brma_eng <- st_transform(sf_brma_eng, crs = 27700) %>% 
  rowid_to_column('row_id')

# # checking 2020 version
# # imports without CRS so set
# sf_brma_eng_20 <- st_read(here("data-input", "brma_eng", "English BRMA(LHA) Layer.gml"))
# # st_crs(sf_brma_eng_20)
# sf_brma_eng_20 <- st_set_crs(sf_brma_eng_20, 27700) %>% 
#   rowid_to_column('row_id')
# 
# # compare - same names in both
# names_14 <- unique(sf_brma_eng$Name)
# names_20 <- unique(sf_brma_eng_20$Name)
# sum(names_14 == names_20)
# 
# # makes area of brma in each and compare
# sf_brma_eng$area14 <- st_area(temp14)
# sf_brma_eng_20$area20 <- st_area(temp20)
# sf_brma_eng_20 %>% 
#   cbind(sf_brma_eng %>% select(area14)) %>% 
#   ggplot(aes(x = as.numeric(area14), y = as.numeric(area20))) +
#   geom_point()

```

### Scotland
Boswarva, Owen. (2017). Broad Rental Market Areas (Scotland), [Dataset]. independent researcher. https://doi.org/10.7488/ds/1968. Obtained using FOI. Open Government Licence. 
Dataset released by Rent Service Scotland (part of the Scottish Government) on 14 October 2014. 

```{r brma scot}

# opena and convert to BNG
sf_brma_scot <- st_read(here("data-input", "brma_scot", "Scotland.shp"))
# st_crs(sf_brma_scot)
sf_brma_scot <- st_transform(sf_brma_scot, crs = 27700) %>% 
  rowid_to_column('row_id')

```

### Wales
Boswarva, Owen. (2017). Broad Rental Market Areas (Wales), [Dataset]. independent researcher. https://doi.org/10.7488/ds/1965. Obtained using FOI. Open Government Licence.
Dataset released by Rent Officers Wales (part of the Housing Division of the Welsh Government) on 23 September 2014.

```{r brma wales}

# imports as BNG
sf_brma_wales <- st_read(here("data-input", "brma_wales", "wales_brma.shp"))
# st_crs(sf_brma_wales)
sf_brma_wales <- st_transform(sf_brma_wales, crs = 27700) %>% 
  rowid_to_column('row_id')


```


## Postcode centroids
Import postcode file with centroids, and split into English, Scottish and Welsh subsets. 
https://geoportal.statistics.gov.uk/datasets/ons::nspl-online-latest-centroids/
Source: Office for National Statistics licensed under the Open Government Licence v.3.0
Contains OS data © Crown copyright and database right [year]

Fields include the following: 

* PCD: 7-character postcodes (may have one space or none)
* PCD2: 8-character postcodes (may have one space or two)
* PCDS: 7- or 8-character postcodes (always has one space)

```{r pcode}

# imports as BNG 
sf_pcode <- st_read(here("data-input", "pcode", "NSPL_Online_Latest_Centroids.shp"))
# st_crs(sf_pcode)

# filter out NI + L9999.. and M9999.. pcodes
sf_pcode <- sf_pcode %>% 
  filter(substr(LAUA, 1, 1) != "N" &
           substr(LAUA, 1, 1) != "L" &
           substr(LAUA, 1, 1) != "M")

# eng
sf_pcode_eng <- sf_pcode[substr(sf_pcode$LAUA, 1, 1) == 'E',] %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, LAUA, OA21, LSOA21) %>% 
  rowid_to_column('row_id')

# scot
sf_pcode_scot <- sf_pcode[substr(sf_pcode$LAUA, 1, 1) == 'S',] %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, LAUA, OA21, LSOA21) %>% 
  rowid_to_column('row_id')

# wales
sf_pcode_wales <- sf_pcode[substr(sf_pcode$LAUA, 1, 1) == 'W',] %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, LAUA, OA21, LSOA21) %>% 
  rowid_to_column('row_id')


# # check
# nrow(sf_pcode) - nrow(sf_pcode_eng) - nrow(sf_pcode_scot) - nrow(sf_pcode_wales)

```


## Pcode in BRMAs
Make the lookup between postcodes and BRMAs, then add variables for each and reduce to those necessary for the lookup file. NB that there are many postcode centroids which fall outside any BRMA - almost entirely coastal. Seem to be using different definitions of the coast. For speed, do most of them using sf_intersects(), then identify the unmatched postcodes, and use sf_nearest() (much slower). This process is why it was necessary to split the postcode centroid file by country first. 

### England
```{r pcode brma eng}

# england 2014 
df_pcode_brma_eng <- as.data.frame(st_intersects(sf_pcode_eng, sf_brma_eng, sparse = TRUE)) %>% 
  left_join(sf_pcode_eng, by = c('row.id' = 'row_id')) %>% 
  left_join(sf_brma_eng, by = c('col.id' = 'row_id')) %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, OA21, LAUA, LSOA21, LOCALITY_I, Name) %>% 
  rename(brma = LOCALITY_I, 
         brma_name = Name)

# sf object with unmatched pcodes
sf_missed <- sf_pcode_eng[!(sf_pcode_eng$PCD %in% df_pcode_brma_eng$PCD), ]

# brma_id for unmatched pcodes  
sf_missed$brma_id <- st_nearest_feature(sf_missed, sf_brma_eng)

# add brma details and make df
df_missed <- as.data.frame(sf_missed) %>% 
  left_join(as.data.frame((sf_brma_eng %>% select(row_id, LOCALITY_I, Name))), 
            by = c('brma_id' = 'row_id')) %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, OA21, LAUA, LSOA21, LOCALITY_I, Name) %>% 
  rename(brma = LOCALITY_I, 
         brma_name = Name)

# merge on to rest
df_pcode_brma_eng <- df_pcode_brma_eng %>% 
  rbind(df_missed)

```


### Scotland
```{r pcode brma scot}

# scotland - OA21/LSOA21 not yet on file
df_pcode_brma_scot <- as.data.frame(st_intersects(sf_pcode_scot, sf_brma_scot, sparse = TRUE)) %>% 
  left_join(sf_pcode_scot, by = c('row.id' = 'row_id')) %>% 
  left_join(sf_brma_scot, by = c('col.id' = 'row_id')) %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, OA21, LAUA, LSOA21, LOCALITY_I, Name) %>% 
  filter(!is.na(PCD)) %>% 
  rename(brma = LOCALITY_I, 
         brma_name = Name)

sum(is.na(df_pcode_brma_scot$PCD))

# sf object with unmatched pcodes
sf_missed <- sf_pcode_scot[!(sf_pcode_scot$PCD %in% df_pcode_brma_scot$PCD), ]

# brma_id for unmatched pcodes  
sf_missed$brma_id <- st_nearest_feature(sf_missed, sf_brma_scot)

# add brma details and make df
df_missed <- as.data.frame(sf_missed) %>% 
  left_join(as.data.frame((sf_brma_scot %>% select(row_id, LOCALITY_I, Name))), 
            by = c('brma_id' = 'row_id')) %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, OA21, LAUA, LSOA21, LOCALITY_I, Name) %>% 
  rename(brma = LOCALITY_I, 
         brma_name = Name)

# merge on to rest
df_pcode_brma_scot <- df_pcode_brma_scot %>% 
  rbind(df_missed)

```


### Wales
```{r pcode brma wales}

# wales - OA11 seems to be missing
df_pcode_brma_wales <- as.data.frame(st_intersects(sf_pcode_wales, sf_brma_wales, sparse = TRUE)) %>% 
  left_join(sf_pcode_wales, by = c('row.id' = 'row_id')) %>% 
  left_join(sf_brma_wales, by = c('col.id' = 'row_id')) %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, OA21, LAUA, LSOA21, brma, brma_name) %>% 
  filter(!is.na(PCD)) 
  

# sf object with unmatched pcodes
sf_missed <- sf_pcode_wales[!(sf_pcode_wales$PCD %in% df_pcode_brma_wales$PCD), ]

# brma_id for unmatched pcodes  
sf_missed$brma_id <- st_nearest_feature(sf_missed, sf_brma_wales)

# add brma details and make df
df_missed <- as.data.frame(sf_missed) %>% 
  left_join(as.data.frame((sf_brma_wales %>% select(row_id, brma, brma_name))), 
            by = c('brma_id' = 'row_id')) %>% 
  select(PCD, PCD2, PCDS, OA11, LSOA11, WZ11, OA21, LAUA, LSOA21, brma, brma_name)

# merge on to rest
df_pcode_brma_wales <- df_pcode_brma_wales %>% 
  rbind(df_missed)

```


## Combine into one data frame

```{r combine}

df_pcode_brma <- df_pcode_brma_eng %>% 
  select(PCD, PCD2, PCDS, brma, brma_name) %>% 
  rbind(df_pcode_brma_scot %>% 
          select(PCD, PCD2, PCDS, brma, brma_name)) %>% 
  rbind(df_pcode_brma_wales %>% 
          select(PCD, PCD2, PCDS, brma, brma_name))

df_pcode_brma <- df_pcode_brma_eng %>% 
  rbind(df_pcode_brma_scot) %>% 
  rbind(df_pcode_brma_wales) 

# # check
# nrow(df_pcode_brma) - nrow(sf_pcode)

```

## Save pcode to brma lookup

```{r save}

# pcode to brma
write_csv((df_pcode_brma %>% 
             select(PCD, PCD2, PCDS, brma, brma_name)), 
          here("data-output", "pcode_brma_lookup.csv"))

```

